Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course

### 2024-08-07 paper/paper.qmd
changed bin width

---
title: "Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course"
author: 
  - Rohan Alexander
  - Luca Carnegie
  - Nathalie Moon
thanks: "Code and data are available at: https://github.com/lcarnegie/llms-achievement. We thank Nathalie Moon for developing and sharing her survey questions. We thank Tiffany Timbers and attendees at JSM 2024 for helpful suggestions. This research is currently under review by the University of Toronto's University Research Ethics Board. Comments can be sent to: rohan.alexander@utoronto.ca."
date: today
date-format: long
abstract: "To help understand the effect of Large Language Models (LLMs) on data science practice we examine the extent to which LLM usage is correlated with the mark that a student gets on a final paper in a classroom data science setting. We find some very mild evidence from this observational study that LLM usage may be associated with better scores, but our main conclusion is that there is no clear relationship between more extensive LLM usage and the student's mark. Despite the classroom setting used for evaluation, the particular activity of interest reflects the work done by professional data scientists. Our finding suggests the need for more extensive work evaluating how LLMs can be integrated into the data science workflow in a way that provides value."
format: pdf
number-sections: true
table-of-contents: false
bibliography: references.bib
---

@@

The goal of our modelling strategy is to better understand how respondents result on their final paper is associated with their self-reported LLM usage. The result on the final paper is a proportion and some respondents got full marks and so we use zero-one-inflated beta regression. Here we briefly describe the model that we use, which follows @kurztotherescue. Diagnostics are included in [Appendix -@sec-model-details].

@@

We estimate the model in `R` [@citeR] using `brms` [@brms].

The expected relationship between LLM usage and final mark is unclear. It may be that stronger students used LLMs in a more sophisticated way or that they did not need to use them. However, the relationship between GPA and final mark is expected to be positive.

>>>>

---
title: "Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course"
author: 
  - Rohan Alexander
  - Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/llms-achievement. We thank Nathalie Moon for developing and sharing her survey questions. We thank Tiffany Timbers and attendees at JSM 2024 for helpful suggestions. This research is currently under review by the University of Toronto's University Research Ethics Board. Comments can be sent to: rohan.alexander@utoronto.ca."
date: today
date-format: long
abstract: "To help understand the effect of Large Language Models (LLMs) on data science practice we examine the extent to which LLM usage is correlated with the mark that a student gets on a final paper in a classroom data science setting. We find some very mild evidence from this observational study that LLM usage may be associated with better scores, but our main conclusion is that there is no clear relationship between more extensive LLM usage and the student's mark. Despite the classroom setting used for evaluation, the particular activity of interest reflects the work done by professional data scientists. Our finding suggests the need for more extensive work evaluating how LLMs can be integrated into the data science workflow in a way that provides value."
format: pdf
number-sections: true
table-of-contents: false
bibliography: references.bib
---

@@

The goal of our modelling strategy is to better understand how a respondents' result on their final paper associates with their self-reported LLM usage.  Students received their result on the final paper as a proportion (a 0-100% mark), so we use zero-one-inflated beta regression to best fit the distribution in this interval. We use this type of regression given the ability of the beta distribution to effectively describe continuous and skewed or heteroskedastic datasets, which we can see in the asymmetrical histogram of the distribution of marks.

```{r final-grade-dist}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Distribution of grades on Final Paper

all_data |>
  select(mark) |>
  ggplot(aes(x = mark)) +
  geom_histogram(binwidth = 0.01) +
  labs(x = "Final paper mark (percentage)", y = "Count") +
  theme_minimal()

```

Transforming the regression with a logit function as the link function ensures that the predicted values stay between the 0-1 range, like in logistic regression. This allows the regression model to only return values between the 0-1, respecting the boundaries of the data. Finally, since students' grades could be either 0, 100, or in between, the zero-one-inflated beta model design is used to include grades that are 0 and 1 as well.

Here we briefly describe the model that we use, which follows @kurztotherescue. Model Diagnostics are included in [Appendix -@sec-model-details].

@@

We estimate the model in `R` [@citeR] as well, using `brms` [@brms].

The expected relationship between LLM usage and final mark is unclear. It could be that stronger students used LLMs in a more sophisticated way or that they did not need to use them. However, the relationship between GPA and final mark is expected to be positive.
